---
title: "Flow Cytometry Analysis — mCitrine Expression in S. cerevisiae"
author: "Alex Gresham"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_notebook: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Background

This analysis examines mCitrine fluorescence data collected on June 9, 2026, using the
Cytek Aurora flow cytometer at the Gresham Laboratory, NYU. Four strains were analyzed:
two negative control strains that lack the mCitrine gene (DGY3283, DGY2807), one
constitutively mCitrine-expressing strain (DGY3955), and one experimental strain of
interest (DGY3954).

mCitrine is a yellow fluorescent protein (excitation ~516 nm, emission ~529 nm) detected
via the B2-A channel. Fluorescence intensity is used as a proxy for mCitrine copy number
and expression level per cell.

The full class dataset (`June9_Scer_students.CSV`) contains gated summary statistics
exported from the Cytek Aurora software. This notebook filters to the samples run by
Alex Gresham (rows 14–17, identified by the suffix "AG").

---

## Load Packages

```{r libraries}
library(tidyverse)
library(RColorBrewer)
```

---

## Load and Filter Data

```{r load-data}
# Read exported Cytek summary statistics
flow_raw <- read_csv("June9_Scer_students.CSV", show_col_types = FALSE)

# The first column contains sample names; rename it
colnames(flow_raw)[1] <- "Sample"

# Filter to Alex Gresham's samples (suffix "AG") — rows 14–17
flow_ag <- flow_raw %>%
  filter(str_ends(Sample, "AG"))

# Extract strain name from sample identifier
flow_ag <- flow_ag %>%
  mutate(Strain = str_extract(Sample, "\\d+(?=AG)"),
         Strain = paste0("DGY", Strain))

# Select key columns
flow_summary <- flow_ag %>%
  select(
    Strain,
    Singlets      = `FSC_Singlets Count`,
    mCit_count    = `mCitrine Count`,
    mCit_pct      = `mCitrine % Parent`,
    Median_B2A    = `FSC_Singlets Median B2-A`
  ) %>%
  mutate(
    mCit_pct   = as.numeric(mCit_pct),
    Median_B2A = as.numeric(Median_B2A),
    Singlets   = as.numeric(Singlets),
    mCit_count = as.numeric(mCit_count)
  )

knitr::kable(flow_summary, digits = 2,
             col.names = c("Strain", "FSC Singlets", "mCitrine+ Cells",
                           "mCitrine % of Singlets", "Median B2-A"))
```

---

## Plot 1 — Percentage of mCitrine-Positive Cells by Strain

```{r bar-chart-pct, fig.width=6, fig.height=4}
# Define strain order and labels
strain_order <- c("DGY3283", "DGY2807", "DGY3955", "DGY3954")
strain_labels <- c("DGY3283\n(0-copy ctrl)", "DGY2807\n(0-copy ctrl)",
                   "DGY3955\n(mCitrine+)", "DGY3954\n(experimental)")

flow_summary %>%
  mutate(Strain = factor(Strain, levels = strain_order)) %>%
  ggplot(aes(x = Strain, y = mCit_pct,
             fill = mCit_pct > 5)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(mCit_pct, 1), "%")),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_x_discrete(labels = strain_labels) +
  scale_fill_manual(values = c("FALSE" = "#aec6cf", "TRUE" = "#2a7f7f"),
                    guide = "none") +
  scale_y_continuous(limits = c(0, 110), expand = c(0, 0)) +
  labs(
    title = "mCitrine Expression by Strain",
    subtitle = "Percentage of FSC-gated singlets in the mCitrine+ gate",
    x = NULL,
    y = "% mCitrine-positive cells"
  ) +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

## Plot 2 — Median B2-A Fluorescence Intensity

The B2-A channel measures mCitrine fluorescence. Comparing median B2-A across
strains gives a sense of per-cell fluorescence intensity, independent of gating.

```{r bar-chart-b2a, fig.width=6, fig.height=4}
flow_summary %>%
  mutate(Strain = factor(Strain, levels = strain_order)) %>%
  ggplot(aes(x = Strain, y = Median_B2A, fill = Median_B2A > 5000)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = formatC(Median_B2A, format = "f", digits = 0, big.mark = ",")),
            vjust = -0.4, size = 3.2) +
  scale_x_discrete(labels = strain_labels) +
  scale_fill_manual(values = c("FALSE" = "#aec6cf", "TRUE" = "#2a7f7f"),
                    guide = "none") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Median B2-A Fluorescence by Strain",
    subtitle = "Median mCitrine channel intensity across all FSC-gated singlets",
    x = NULL,
    y = "Median B2-A (a.u.)"
  ) +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

## Plot 3 — Context: Alex's Strains vs. Full Class Dataset

Compare mCitrine% across all 17 samples to place Alex's results in context.

```{r full-context, fig.width=10, fig.height=5}
flow_raw_clean <- flow_raw %>%
  rename(Sample = 1) %>%
  mutate(
    mCit_pct = as.numeric(`mCitrine % Parent`),
    Analyst  = case_when(
      str_ends(Sample, "AG")  ~ "Alex Gresham",
      str_ends(Sample, "LCS") ~ "Other student",
      str_ends(Sample, "TJ")  ~ "Other student",
      TRUE                    ~ "Other student"
    ),
    Short = str_remove(Sample, "Students_June9-")
  )

ggplot(flow_raw_clean, aes(x = reorder(Short, mCit_pct),
                            y = mCit_pct, fill = Analyst)) +
  geom_col() +
  scale_fill_manual(values = c("Alex Gresham" = "#2a7f7f",
                               "Other student" = "#b0bec5")) +
  coord_flip() +
  labs(
    title = "mCitrine Expression — Full Class Dataset (June 9, 2026)",
    x = NULL, y = "% mCitrine-positive cells", fill = NULL
  ) +
  theme_classic(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

---

## Summary and Biological Interpretation

```{r summary-table}
flow_summary %>%
  mutate(
    Interpretation = case_when(
      mCit_pct < 1   ~ "Negative control — no mCitrine gene present",
      mCit_pct > 95  ~ "Constitutive mCitrine expression — single copy integration",
      TRUE           ~ "Heterogeneous expression — possible CNV or regulated promoter"
    )
  ) %>%
  knitr::kable(digits = 2,
               col.names = c("Strain", "Singlets", "mCitrine+ Cells",
                             "mCitrine%", "Median B2-A", "Interpretation"))
```

**Key findings:**

- **DGY3283 and DGY2807** serve as zero-copy negative controls. Both show <0.05% mCitrine-positive cells, confirming that the gate effectively excludes non-fluorescent cells.

- **DGY3955** shows 99.69% mCitrine positivity, consistent with constitutive single-copy expression at a stable genomic locus (e.g., a neutral locus integration). Virtually all singlet cells fall within the mCitrine gate.

- **DGY3954** is the most biologically interesting sample: 42.79% of singlet cells are mCitrine-positive while 57.21% are not, producing a **heterogeneous population**. In the context of GAP1 copy number variation reporter strains, this bimodal distribution is consistent with an active CNV event, where a subpopulation of cells has amplified the GAP1-mCitrine locus in response to nutrient stress, while the remainder retain the ancestral copy number. Alternatively, the mCitrine gene may be under a regulated promoter active in only a subset of cells.

---

*Analysis conducted June 2026 · Gresham Lab, New York University*
